---
title: "Gov 50 Final Project"
author: "Sara Kumar"
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---

## Project thoughts
I am interested in exploring data related to US House Representatives and the average ages of their respective constituencies. I also would like to explore the differences between the ages of State House Representatives and Federal Representatives. I would like to explore what ages are less represented in this branch of government at each level.

## Milestone 2
Does political affiliation influence the race of female elected officials in Congress (specifically the House of Representatives)? In this study, I plan to examine race trends in female elected representatives in Congress to see if greater numbers of POC female representatives relate to party affiliation. I hypothesize that Democrats will demonstrate greater POC female representation in their elected representatives. As seen historically, California, a left-leaning state, has elected more African Americans to Congress than any other state. Furthermore, Congress has been getting progressively more diverse in recent years. The explanatory variable of interest is the amount of Democratic POC female representatives and the amount of Republican POC female representatives. These two explanatory variables will then be compared in external analyses. This variable will be measured from the Center for American Women and Politics’ data set on elected House of Representative officials. If I observe greater numbers of POC female elected officials in Democratic parties relative to Republicans, this would provide support for my hypothesis. If, on the other hand, I observe less POC female representation in left-leaning states as compared to right, this would provide evidence against my hypothesis.

```{r}
library(tidyverse)
```


```{r}
stateofficials <- read.csv("officeholders_positions.csv")

stateofficials
```
```{r}
dems <- stateofficials %>%
  filter(party == 'Democrat') 
ggplot(data=dems, aes(x=race_ethnicity)) + 
geom_bar() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5)) +
  labs(title = "Female Democratic Representatives by Race")

dems
```

```{r}
reps <- stateofficials %>%
  filter(party == 'Republican') 
ggplot(data=reps, aes(x=race_ethnicity)) + 
geom_bar() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5)) +
  labs(title = "Female Republican Representatives by Race")

reps
```
## Results
I hypothesized that there would be greater numbers of POC female elected officials in Democratic parties relative to Republican. As seen from the tables above, there are around 6 POC races represented in Democratic POC female elected officials whereas Republican POC female elected officials only demonstrate about 2 POC races. Furthermore, relatively speaking, POC female elected Republican representatives do not reach even half that of the white representatives numbers wise. In contrast, the numer of non-white Democratic representatives is just a little less than the amount of white elected officials. Overall, my findings seem to prove my hypothesis.