---
title: "Gov 50 Final Project"
author: "Sara Kumar"
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---

## Project thoughts
I am interested in exploring data related to US House Representatives and the average ages of their respective constituencies. I also would like to explore the differences between the ages of State House Representatives and Federal Representatives. I would like to explore what ages are less represented in this branch of government at each level.

## Milestone 2
Does the political affiliation of the state influence the number of female elected officials in Congress? In this study, I plan to examine gender trends in elected officials in Congress to see if first, there are greater numbers of female officials in certain states, and second, if this trend relates to the party affiliation of that region. I hypothesize that more Democratic-leaning states will have larger numbers of female senators/representatives. As seen historically, California, a left-leaning state, has elected more women to Congress than any other state. Furthermore, Congress has been getting progressively more female in recent years. The explanatory variable of interest is what political affiliation each state is. The variable is coded = 1 for Democratic and =0 for Republican. My outcome variable is the count of female elected officials in Congress for each respective state. For instance, if the variable is coded = 13, the data is indicating that there are 13 female representatives in Congress from that state. This variable will be measured from the Center for American Women and Politics’ dataset on Women elected officials. If I observe greater numbers of female elected officials in Democratic states relative to Republicans, this would provide support for my hypothesis. If, on the other hand, I observe less female representation in left-leaning states as compared to right, this would provide evidence against my hypothesis.

```{r}
library(tidyverse)
```


```{r}
stateofficials <- read.csv("stateofficials.csv")

stateofficials
```
```{r}
stateofficials |>
  group_by(state, party) |>
  ggplot(mapping = aes(x = party, y = state)) +
  geom_point()
```


